---
title: "aotus.glmms"
author: "Isabel Arcusa"
date: "2024-08-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The goal of this script is to generate generalized linear mixed models (GLMM's) for response behaviors observed in the Aotus 2021 experiment. The models will be run using glmmTMB().  

The code written here will use glmmTMB() to generate GLMM's for ten response behaviors (arch, subcaudal scent marking, anogenital sniffing, urine drinking, close proximity, Hinde's index, attempts to mate/mount, and defecating) in four parts:  
1) Discuss justification for using glmmTMB()     
2) Import packages and data  
3) Prepare data frames required for models
4) Make models

**Part 1: Using glmmTMB()**  
The decision to use glmmTMB() was made from the following points:  
A) Some of the response behaviors we are interested in modeling are zero-inflated and/or beta-distributed.  
B) In the first round of GLMM's, many of the models made with glmer() struggled to converge. The warnings produced with those models' outcomes suggest that the models were either overfit with too many random effects or that the data's distribution is not best modeled by glmer().  
C) Another recent study by Miaretsoa et al. with similar statistical analyses uses glmmTMB().  
D) Results from glmmTMB() can be used in model averaging, which is helpful when looking for the best fitting models.  

**Part 2: Import packages and data**  
```{r}
# Import packages
library(dplyr)      # dplyr for data manipulation
library(ggplot2)    # ggplot2 for data visualization
library(glmmTMB)    # glmmTMB for generating mixed effect linear models 
```

```{r}
# Set working directory
setwd("~/Desktop/aotus.2024")

# Import raw experiment data
raw.df<-read.csv("chemical.mateguarding.data.csv", header = TRUE)

# Eliminate columns that are not needed for these models
# Rename edited data frame as df
df<-raw.df %>% select(-uw.and.agscratch, -prx.23, -anointing, -chuck, -yawn, -lip.smack, 
                      -self.groom, -partner.groom)

# Round chirp and urine columns to the nearest whole number and add as a new column
# We can use mutate() and round() from dplyr() to achieve this
# Add a rounded chirp column
df<-df %>%
  mutate(chirp.rounded = round(df$chirp))

# Add a rounded urine column
df<-df %>%
  mutate(urine.rounded = round(df$urine))

# Some values for prx.01 will have to be rounded
# Add a rounded prx.01 column
df<-df %>%
  mutate(prx.01.rounded = round(df$prx.01))

# Verify that the data frame looks as it should
head(df)
```

**Part 3: Prepare data frames**  
Since we know through observation that males and females display some behaviors at different rates, we can further refine some models by taking out sex as a factor. We can achieve this by creating two separate data frames: one for males and one for females. We can then run two separate models to determine the effect of treatment on urination for each sex. This will also allow us to examine the effect of contraceptive status in each sex without running into issues of "None" entries applying to both males and females.    

```{r}
# Remove rows where sex is "Male"
df.female <- df %>%
  filter(df$sex != "Male")

# Check to make sure the data frame only has female entries
head(df.female)

# Remove rows where sex is "Female"
df.male <- df %>%
  filter(df$sex != "Female")

# Check to make sure the data frame only has male entries
head(df.male)
```

**Part 4: Make models**  
```{r}
# Define all variables
name<-df$name
cage<-df$cage
sex<-df$sex
treatment<-df$treatment
contraceptive.status<-df$contraceptive.status
arch<-df$arch
chirp.rounded<-df$chirp.rounded
urine.rounded<-df$urine.rounded
subc.sm<-df$subc.sm
ags<-df$ags
ud<-df$ud
atm.mount<-df$atm.mount
defecate<-df$defecate
prx.01.rounded<-df$prx.01.rounded
hindes.index<-df$hindes.index
```

```{r}
# To ensure that the model output compares the significance of treatment to the
# control, we have to specify the reference category
df$treatment <- factor(df$treatment, levels = c("Saline.Control", "Female.Urine", "Male.Urine"))
df$treatment <- relevel(df$treatment, ref = "Saline.Control")

df$contraceptive.status <- factor(df$contraceptive.status, levels = c("Contraception", "None", "Vasectomized"))
df$contraceptive.status <- relevel(df$contraceptive.status, ref = "None")
```

Many of our response behaviors are negative binomial-distributed response variables that can be zero. Therefore the family argument should be set to negative binomial (see https://glmmtmb.github.io/glmmTMB/articles/glmmTMB.html for more info). Defining the family argument as negative binomial comes with two options: nbinom1 or nbinom2 with a log link. nbinom1 is better for when the mean and variance have a linear relationship. nbinom2 is better for when variance is a quadratic function of the mean. To determine the best fit, both versions will be used. Then, the AIC values from both models will be compared.  

**Arch**  
```{r}
# Make the models using glmmTMB package
# name and cage are random effects
# sex, treatment, and contraceptive.status are fixed effects

# nbinom1
arch.model.1<-glmmTMB(arch ~ sex + sex*treatment + (1|name) + (1|cage), data=df, ziformula=~1, family=nbinom1(link = "log"))

summary(arch.model.1)

# nbinom2
arch.model.2<-glmmTMB(arch ~ sex + sex*treatment + (1|name) + (1|cage), data=df, ziformula=~1, family=nbinom2(link = "log"))

summary(arch.model.2)

# Compare AIC's from arch.model.1 and arch.model.2
AIC(arch.model.1, arch.model.2)
```
arch.model.1 has a slightly lower AIC, and thus is the better fitting model.  

*Male Arch*  
```{r}
# Run model including contraceptive.status as a fixed effect
male.arch.model.1<-glmmTMB(arch ~ treatment + contraceptive.status + treatment*contraceptive.status + (1|name) + (1|cage), data=df.male, ziformula=~1, family=nbinom1(link = "log"))

summary(male.arch.model.1)

# We have an issue with model convergence as evidenced by the NA values
# Remove random effects one at a time
male.arch.model.2<-glmmTMB(arch ~ treatment + contraceptive.status + treatment*contraceptive.status + (1|name), data=df.male, ziformula=~1, family=nbinom1(link = "log"))

summary(male.arch.model.2)

# Remove name
male.arch.model.3<-glmmTMB(arch ~ treatment + contraceptive.status + treatment*contraceptive.status, data=df.male, ziformula=~1, family=nbinom1(link = "log"))

summary(male.arch.model.3)
```

*Female Arch*  
```{r}
# Run model including contraceptive.status as a fixed effect
female.arch.model<-glmmTMB(arch ~ treatment, data=df.female, ziformula=~1, family=nbinom1(link = "log"))

summary(female.arch.model)
```
 
**Chirp**  
```{r}
# chirp.rounded is a negative binomial-distributed response variable that can be zero
# therefore family should be set to negative binomial

# nbinom1
chirp.model.1<-glmmTMB(chirp.rounded ~ sex + sex*treatment + (1|name) + (1|cage), data=df, ziformula=~1, family=nbinom1(link = "log"))

summary(chirp.model.1)

# nbinom2
chirp.model.2<-glmmTMB(chirp.rounded ~ sex + sex*treatment + (1|name) + (1|cage), data=df, ziformula=~1, family=nbinom2(link = "log"))

summary(chirp.model.2)

# Compare AIC's from chirp.model.1 and chirp.model.2
AIC(chirp.model.1, chirp.model.2)
```

*Male Chirp*  
```{r}
# Run model including contraceptive.status as a fixed effect
male.chirp.model<-glmmTMB(chirp.rounded ~ treatment + contraceptive.status + treatment*contraceptive.status + (1|name) + (1|cage), data=df.male, ziformula=~1, family=nbinom1(link = "log"))

summary(male.chirp.model)
```

*Female Chirp*  
```{r}
# Run model including contraceptive.status as a fixed effect
female.chirp.model<-glmmTMB(chirp.rounded ~ treatment + contraceptive.status + treatment*contraceptive.status + (1|name) + (1|cage), data=df.female, ziformula=~1, family=nbinom1(link = "log"))

summary(female.chirp.model)
```

**Urine**  
```{r}
# urine.rounded is a negative binomial-distributed response variable that can be zero
# therefore family should be set to negative binomial

# nbinom1
urine.model.1<-glmmTMB(urine.rounded ~ sex + sex*treatment + (1|name) + (1|cage), data=df, ziformula=~1, family=nbinom1(link = "log"))

summary(urine.model.1)

# nbinom2
urine.model.2<-glmmTMB(urine.rounded ~ sex + sex*treatment + (1|name) + (1|cage), data=df, ziformula=~1, family=nbinom2(link = "log"))

summary(urine.model.2)

# Compare AIC's from urine.model.1 and urine.model.2
AIC(urine.model.1, urine.model.2)
```

*Male Urine*  
```{r}
# Male only urine model
male.urine.model<-glmmTMB(urine.rounded ~ treatment + contraceptive.status + treatment*contraceptive.status + (1|name) + (1|cage), data=df.male, ziformula=~1, family=nbinom1(link = "log"))

summary(male.urine.model)
```

*Female Urine*  
```{r}
# Female only urine model
female.urine.model.1<-glmmTMB(urine.rounded ~ treatment + contraceptive.status + treatment*contraceptive.status + (1|name) + (1|cage), data=df.female, ziformula=~1, family=nbinom1(link = "log"))

summary(female.urine.model.1)

# Remove cage
female.urine.model.2<-glmmTMB(urine.rounded ~ treatment + contraceptive.status + treatment*contraceptive.status + (1|name), data=df.female, ziformula=~1, family=nbinom1(link = "log"))

summary(female.urine.model.2)

# Remove name
female.urine.model.3<-glmmTMB(urine.rounded ~ treatment + contraceptive.status + treatment*contraceptive.status, data=df.female, ziformula=~1, family=nbinom1(link = "log"))

summary(female.urine.model.3)
```

**Subcaudal Scent Mark**  
```{r}
# subc.sm is a negative binomial-distributed response variable that can be zero
# therefore family should be set to negative binomial

# nbinom1
subc.sm.model.1<-glmmTMB(subc.sm ~ sex + sex*treatment + (1|name) + (1|cage), data=df, ziformula=~1, family=nbinom1(link = "log"))

summary(subc.sm.model.1)

# nbinom2
subc.sm.model.2<-glmmTMB(subc.sm ~ sex + sex*treatment + (1|name) + (1|cage), data=df, ziformula=~1, family=nbinom2(link = "log"))

summary(subc.sm.model.2)

# Compare AIC's from subc.sm.model.1 and subc.sm.model.2
AIC(subc.sm.model.1, subc.sm.model.2)
```
subc.sm.model.2 has a lower AIC, and thus is the better fitting model.  

*Male Scent mark*  
```{r}
# Male only subc.sm model
male.subcsm.model<-glmmTMB(subc.sm ~ treatment + contraceptive.status + treatment*contraceptive.status + (1|name) + (1|cage), data=df.male, ziformula=~1, family=nbinom1(link = "log"))

summary(male.subcsm.model)
```

*Female Scent mark*  
```{r}
# Female only subc.sm model
female.subcsm.model.1<-glmmTMB(subc.sm ~ treatment + contraceptive.status + treatment*contraceptive.status + (1|name) + (1|cage), data=df.female, ziformula=~1, family=nbinom1(link = "log"))

summary(female.subcsm.model.1)

# Remove cage
female.subcsm.model.2<-glmmTMB(subc.sm ~ treatment + contraceptive.status + treatment*contraceptive.status + (1|name), data=df.female, ziformula=~1, family=nbinom1(link = "log"))

summary(female.subcsm.model.2)
```

**Anogenital Sniff**  
```{r}
# ags is a negative binomial-distributed response variable that can be zero
# therefore family should be set to negative binomial

# nbinom1
ags.model.1<-glmmTMB(ags ~ sex + sex*treatment + (1|name) + (1|cage), data=df, ziformula=~1, family=nbinom1(link = "log"))

summary(ags.model.1)

# nbinom2
ags.model.2<-glmmTMB(ags ~ sex + sex*treatment + (1|name) + (1|cage), data=df, ziformula=~1, family=nbinom2(link = "log"))

summary(ags.model.2)

# There is an issue with model convergence with the above models
# The most likely reason is that there is some issue with the variables
# Either there are too many variables or some collinearity
# To solve this, let's re-run ags.model.1 removing the random variable, cage

ags.model.1b<-glmmTMB(ags ~ sex + sex*treatment + (1|name), data=df, ziformula=~1, family=nbinom1(link = "log"))

summary(ags.model.1b)

# No more errors! The model ags.model.1b can converge
# Repeat with ags.model.2

ags.model.2b<-glmmTMB(ags ~ sex + sex*treatment + (1|name), data=df, ziformula=~1, family=nbinom2(link = "log"))

summary(ags.model.2b)

# Compare AIC's from ags.model.1 and ags.model.2
AIC(ags.model.1b, ags.model.2b)
```
ags.model.1b has a lower AIC, and thus is the better fitting model.  

*Male Anogenital sniff*  
```{r}
male.ags.model<-glmmTMB(ags ~ treatment + contraceptive.status + treatment*contraceptive.status + (1|name) + (1|cage), data=df.male, ziformula=~1, family=nbinom1(link = "log"))

summary(male.ags.model)
```

**Urine drink**  
```{r}
# ud is a negative binomial-distributed response variable that can be zero
# therefore family should be set to negative binomial

# nbinom1
ud.model.1<-glmmTMB(ud ~ sex + sex*treatment + (1|name) + (1|cage), data=df, ziformula=~1, family=nbinom1(link = "log"))

summary(ud.model.1)

# nbinom2
ud.model.2<-glmmTMB(ud ~ sex + sex*treatment + (1|name) + (1|cage), data=df, ziformula=~1, family=nbinom2(link = "log"))

summary(ud.model.2)

# Once again, we have a model convergence issue
# Let's re-run ud.model.1 removing the random variable, cage

ud.model.1b<-glmmTMB(ud ~ sex + sex*treatment + (1|name), data=df, ziformula=~1, family=nbinom1(link = "log"))

summary(ud.model.1b)

# The model atm.mount.model.1b still cannot converge
# Remove name this time
ud.model.1c<-glmmTMB(ud ~ sex + sex*treatment, data=df, ziformula=~1, family=nbinom1(link = "log"))

summary(ud.model.1c)

# Remove sex*treatment interaction and use treatment
ud.model.1d<-glmmTMB(ud ~ sex + treatment, data=df, ziformula=~1, family=nbinom1(link = "log"))

summary(ud.model.1d)

# No NA's and no issues with convergence
# However, we get z values of 0.000 and Pr(>z) of 1.000
# We're still having an issue with collinearity
# Let's try removing sex, since we know only males urine drink
ud.model.1e<-glmmTMB(ud ~ treatment, data=df, ziformula=~1, family=nbinom1(link = "log"))

summary(ud.model.1e)

# Much better! Repeat with ud.model.2
ud.model.2e<-glmmTMB(ud ~ treatment, data=df, ziformula=~1, family=nbinom2(link = "log"))

summary(ud.model.2e)

# Compare AIC's from ud.model.1 and ud.model.2
AIC(ud.model.1e, ud.model.2e)
```
ud.model.2e has a lower AIC, so it is a better fit than ud.model.1e.  

*Male Urine drink*  
```{r}
male.ud.model.1<-glmmTMB(ud ~ treatment + contraceptive.status + treatment*contraceptive.status + (1|name) + (1|cage), data=df.male, ziformula=~1, family=nbinom1(link = "log"))

summary(male.ud.model.1)

# Remove cage
male.ud.model.2<-glmmTMB(ud ~ treatment + contraceptive.status + treatment*contraceptive.status + (1|name), data=df.male, ziformula=~1, family=nbinom1(link = "log"))

summary(male.ud.model.2)

# Remove name
male.ud.model.3<-glmmTMB(ud ~ treatment + contraceptive.status + treatment*contraceptive.status, data=df.male, ziformula=~1, family=nbinom1(link = "log"))

summary(male.ud.model.3)

# Remove interaction
male.ud.model.4<-glmmTMB(ud ~ treatment + contraceptive.status, data=df.male, ziformula=~1, family=nbinom1(link = "log"))

summary(male.ud.model.4)
```

**Mount attempt**  
```{r}
# atm.mount is a negative binomial-distributed response variable that can be zero
# therefore family should be set to negative binomial

# nbinom1
atm.mount.model.1<-glmmTMB(atm.mount ~ sex + sex*treatment + (1|name) + (1|cage), data=df, ziformula=~1, family=nbinom1(link = "log"))

summary(atm.mount.model.1)

# nbinom2
atm.mount.model.2<-glmmTMB(atm.mount ~ sex + sex*treatment + (1|name) + (1|cage), data=df, ziformula=~1, family=nbinom2(link = "log"))

summary(atm.mount.model.2)

# Once again, we have a model convergence issue
# Let's re-run ud.model.1 removing the random variable, cage

atm.mount.model.1b<-glmmTMB(atm.mount ~ sex + sex*treatment + (1|name), data=df, ziformula=~1, family=nbinom1(link = "log"))

summary(atm.mount.model.1b)

# The model atm.mount.model.1b still cannot converge
# Remove name this time
atm.mount.model.1c<-glmmTMB(atm.mount ~ sex + sex*treatment, data=df, ziformula=~1, family=nbinom1(link = "log"))

summary(atm.mount.model.1c)

# Remove sex*treatment interaction and use treatment
atm.mount.model.1d<-glmmTMB(ud ~ sex + treatment, data=df, ziformula=~1, family=nbinom1(link = "log"))

summary(atm.mount.model.1d)

# No NA's and no issues with convergence
# However, we get z values of 0.000 and Pr(>z) of 1.000
# We're still having an issue with collinearity
# Let's try removing sex, since we know only males really attempt to mate
atm.mount.model.1e<-glmmTMB(atm.mount ~ treatment, data=df, ziformula=~1, family=nbinom1(link = "log"))

summary(atm.mount.model.1e)

# Much better! Repeat with atm.mount.model.2
atm.mount.model.2e<-glmmTMB(atm.mount ~ treatment, data=df, ziformula=~1, family=nbinom2(link = "log"))

summary(atm.mount.model.2e)

# Compare AIC's from atm.mount.model.1 and atm.mount.model.2
AIC(atm.mount.model.1e, atm.mount.model.2e)
```
Both models have the same AIC, so atm.mount.model.1e and atm.mount.model.2e fit the same.  

*Male Attempt to mount*  
```{r}
male.atm.model.1<-glmmTMB(atm.mount ~ treatment + contraceptive.status + treatment*contraceptive.status + (1|name) + (1|cage), data=df.male, ziformula=~1, family=nbinom1(link = "log"))

summary(male.atm.model.1)

# Remove cage
male.atm.model.2<-glmmTMB(atm.mount ~ treatment + contraceptive.status + treatment*contraceptive.status + (1|name), data=df.male, ziformula=~1, family=nbinom1(link = "log"))

summary(male.atm.model.2)

# Remove name
male.atm.model.3<-glmmTMB(atm.mount ~ treatment + contraceptive.status + treatment*contraceptive.status, data=df.male, ziformula=~1, family=nbinom1(link = "log"))

summary(male.atm.model.3)

# Remove interaction
male.atm.model.4<-glmmTMB(atm.mount ~ treatment + contraceptive.status, data=df.male, ziformula=~1, family=nbinom1(link = "log"))

summary(male.atm.model.4)

# Remove treatment
male.atm.model.5<-glmmTMB(atm.mount ~ treatment, data=df.male, ziformula=~1, family=nbinom1(link = "log"))

summary(male.atm.model.5)

# No model convergence was achieved
```

**Defecate**  
```{r}
# defecate is a negative binomial-distributed response variable that can be zero
# therefore family should be set to negative binomial

# nbinom1
defecate.model.1<-glmmTMB(defecate ~ sex + sex*treatment + (1|name) + (1|cage), data=df, ziformula=~1, family=nbinom1(link = "log"))

summary(defecate.model.1)

# nbinom2
defecate.model.2<-glmmTMB(defecate ~ sex + sex*treatment + (1|name) + (1|cage), data=df, ziformula=~1, family=nbinom2(link = "log"))

summary(defecate.model.2)

# Once again, we have a model convergence issue
# Let's re-run defecate.model.1 removing the random variable, cage

defecate.model.1b<-glmmTMB(defecate ~ sex + sex*treatment + (1|name), data=df, ziformula=~1, family=nbinom1(link = "log"))

summary(defecate.model.1b)

# The model atm.mount.model.1b still cannot converge
# Remove name this time
defecate.model.1c<-glmmTMB(defecate ~ sex + sex*treatment, data=df, ziformula=~1, family=nbinom1(link = "log"))

summary(defecate.model.1c)

# Much better! Repeat with atm.mount.model.2
defecate.model.2c<-glmmTMB(defecate ~ sex + sex*treatment, data=df, ziformula=~1, family=nbinom2(link = "log"))

summary(defecate.model.2c)

# Compare AIC's from atm.mount.model.1 and atm.mount.model.2
AIC(defecate.model.1c, defecate.model.2c)
```
defecate.model.1c has a slightly lower AIC than defecate.model.2c, so it is a better fit.    

*Male Defecate*  
```{r}
male.defecate.model.1<-glmmTMB(defecate ~ treatment + contraceptive.status + treatment*contraceptive.status + (1|name) + (1|cage), data=df.male, ziformula=~1, family=nbinom1(link = "log"))

summary(male.defecate.model.1)

# Remove cage
male.defecate.model.2<-glmmTMB(defecate ~ treatment + contraceptive.status + treatment*contraceptive.status + (1|name), data=df.male, ziformula=~1, family=nbinom1(link = "log"))

summary(male.defecate.model.2)

# Remove name
male.defecate.model.3<-glmmTMB(defecate ~ treatment + contraceptive.status + treatment*contraceptive.status, data=df.male, ziformula=~1, family=nbinom1(link = "log"))

summary(male.defecate.model.3)

# Remove interaction
male.defecate.model.4<-glmmTMB(defecate ~ treatment + contraceptive.status, data=df.male, ziformula=~1, family=nbinom1(link = "log"))

summary(male.defecate.model.4)
```

*Female Defecate*  
```{r}
female.defecate.model.1<-glmmTMB(defecate ~ treatment + contraceptive.status + treatment*contraceptive.status + (1|name) + (1|cage), data=df.female, ziformula=~1, family=nbinom1(link = "log"))

summary(female.defecate.model.1)

# Remove cage
female.defecate.model.2<-glmmTMB(defecate ~ treatment + contraceptive.status + treatment*contraceptive.status + (1|name), data=df.female, ziformula=~1, family=nbinom1(link = "log"))

summary(female.defecate.model.2)

# Remove name
female.defecate.model.3<-glmmTMB(defecate ~ treatment + contraceptive.status + treatment*contraceptive.status, data=df.female, ziformula=~1, family=nbinom1(link = "log"))

summary(female.defecate.model.3)
```

*Close Proximity*  

Since close proximity is expected to be the same for males and females scored in any given trial, we can model the effect of treatment on close proximity using only one sex's data.  

```{r}
# nbinom1
prx.01.rounded.model.1<-glmmTMB(prx.01.rounded ~ treatment + (1|name) + (1|cage), data=df.male, ziformula=~1, family=nbinom1(link = "log"))

summary(prx.01.rounded.model.1)

# nbinom2
prx.01.rounded.model.2<-glmmTMB(prx.01.rounded ~ treatment + (1|name) + (1|cage), data=df.male, ziformula=~1, family=nbinom2(link = "log"))

summary(prx.01.rounded.model.2)

# Compare AIC's from prx.01.rounded.model.1 and prx.01.rounded.model.2
AIC(prx.01.rounded.model.1, prx.01.rounded.model.2)
```

*Hinde's Index*  

```{r}
# hindes.index is a normally-distributed response variable

hindes.index.model.1<-glmmTMB(hindes.index ~ treatment + (1|name) + (1|cage), data=df.male, ziformula=~1, family=gaussian())

summary(hindes.index.model.1)

# sexMale z value of 0.000 indicates colinearity
# Remove sex and only examine treatment

hindes.index.model.1b<-glmmTMB(hindes.index ~ treatment + (1|name) + (1|cage), data=df.male, ziformula=~1, family=gaussian())

summary(hindes.index.model.1b)
```