---
title: "aotus.split.violins"
author: "Isabel Arcusa"
date: "2024-08-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This file contains code to generate split violin plots for behaviors in response to treatments. The plots show data from the 2021 chemical mate-guarding Aotus experiment. The violin plots can be made in two ways (both are used to demonstrate their utility):    
1) Show entire male and female distributions side-by-side.  
2) Split distributions by sex and add bars to show means for each sex's distribution.  

For the purpose of this file, we will be using the second method. This method is inspired by Tung on StackOverflow at https://stackoverflow.com/questions/51228076/ggplot-split-violin-plot-with-horizontal-mean-lines.  

Some plots will show differences in behavior between males and females, and others will show the differences between vasectomized and non-vasectomized males. For all plots, the x-axis is the treatment (female urine, male urine, or saline control) and the y-axis is the scale of the behavior.  

```{r}
# Install ggplot2 for graphs and dplyr 
library(ggplot2)
library(dplyr)
library(plyr)

# Set working directory
setwd("~/Desktop/aotus.2024")

# Import raw experiment data
raw.df<-read.csv("chemical.mateguarding.data.csv", header = TRUE)

# Eliminate columns that are not needed for these models
# Rename edited data frame as df
df<-raw.df %>% select(-uw.and.agscratch, -prx.23, -anointing, -chuck, -yawn, -lip.smack, 
                      -self.groom, -partner.groom)

# Round chirp and urine columns to the nearest whole number and add as a new column
# We can use mutate() and round() from dplyr() to achieve this
# Add a rounded chirp column
df<-df %>%
  mutate(chirp.rounded = round(df$chirp))

# Add a rounded urine column
df<-df %>%
  mutate(urine.rounded = round(df$urine))

# Some values for prx.01 will have to be rounded
# Add a rounded prx.01 column
df<-df %>%
  mutate(prx.01.rounded = round(df$prx.01))

# Verify that the data frame looks as it should
head(df)
```

```{r}
# Define all variables
name<-df$name
cage<-df$cage
sex<-df$sex
treatment<-df$treatment
contraceptive.status<-df$contraceptive.status
arch<-df$arch
chirp.rounded<-df$chirp.rounded
urine.rounded<-df$urine.rounded
subc.sm<-df$subc.sm
ags<-df$ags
ud<-df$ud
atm.mount<-df$atm.mount
defecate<-df$defecate
partner.mark<-df$partner.mark
prx.01.rounded<-df$prx.01.rounded
```


```{r}
# Remove rows where contraceptive status is "Vasectomized"
df.none.vas <- df.male %>%
  filter(df.male$contraceptive.status != "Vasectomized")

# Check to make sure the data frame only has vasectomized entries
head(df.none.vas)

# Remove rows where contraceptive status is "None"
df.vas <- df.male %>%
  filter(df.male$contraceptive.status != "None")

# Check to make sure the data frame only has none entries
head(df.vas)
```

The critical function in generating these kinds of split violin plots is called geom_split_violin().  

```{r}
# Define geom_split_violin()
# This is the function that splits the distributions in half by whatever you want to compare
# In our case, it lets us split by sex so we can compare males vs. females

GeomSplitViolin <- ggproto(
  "GeomSplitViolin",
  GeomViolin,
  draw_group = function(self, data, ..., draw_quantiles = NULL) {
    data <- transform(data,
                      xminv = x - violinwidth * (x - xmin),
                      xmaxv = x + violinwidth * (xmax - x)
    )
    grp <- data[1, "group"]
    newdata <- plyr::arrange(
      transform(data, x = if (grp %% 2 == 1) xminv else xmaxv),
      if (grp %% 2 == 1) y else -y
    )
    newdata <- rbind(newdata[1, ], newdata, newdata[nrow(newdata), ], newdata[1, ])
    newdata[c(1, nrow(newdata) - 1, nrow(newdata)), "x"] <- round(newdata[1, "x"])
    if (length(draw_quantiles) > 0 & !scales::zero_range(range(data$y))) {
      stopifnot(all(draw_quantiles >= 0), all(draw_quantiles <= 1))
      quantiles <- ggplot2:::create_quantile_segment_frame(data, draw_quantiles)
      aesthetics <- data[rep(1, nrow(quantiles)), setdiff(names(data), c("x", "y")), drop = FALSE]
      aesthetics$alpha <- rep(1, nrow(quantiles))
      both <- cbind(quantiles, aesthetics)
      quantile_grob <- GeomPath$draw_panel(both, ...)
      ggplot2:::ggname(
        "geom_split_violin",
        grid::grobTree(GeomPolygon$draw_panel(newdata, ...), quantile_grob)
      )
    } else {
      ggplot2:::ggname("geom_split_violin", GeomPolygon$draw_panel(newdata, ...))
    }
  }
)

geom_split_violin <- function(mapping = NULL,
                              data = NULL,
                              stat = "ydensity",
                              position = "identity", ...,
                              draw_quantiles = NULL,
                              trim = TRUE,
                              scale = "area",
                              na.rm = FALSE,
                              show.legend = NA,
                              inherit.aes = TRUE) {
  layer(
    data = data,
    mapping = mapping,
    stat = stat,
    geom = GeomSplitViolin,
    position = position,
    show.legend = show.legend,
    inherit.aes = inherit.aes,
    params = list(
      trim = trim,
      scale = scale,
      draw_quantiles = draw_quantiles,
      na.rm = na.rm, ...
    )
  )
}
```

Now that we've written geom_split_violin(), we can use it to make split violin plots.  

*Arch*  
```{r}
# ggplot(data frame, aes(x variable, y variable, fill = as.factor(splitting category)))

arch.plot<-ggplot(df, aes(treatment, arch, fill = as.factor(sex))) +
  geom_split_violin(trim = TRUE) +
  # stat_summary() lets us display a descriptive statistic on a distribution
  stat_summary(fun = mean, fun.min = mean, fun.max = mean,
               geom = "crossbar", 
               width = 0.25,
               position = position_dodge(width = .25),
  ) +
  labs(x = "Treatment", y = "Arch (intervals/trial)") +
  theme_classic() +
  theme(text = element_text(size = 20)) +
  # scale_x_discrete is our different treatments or categories
  scale_x_discrete(labels = c("0" = "Female_Urine", "1" = "Male_Urine", "2" = "Saline_Control")) +
  # scale_fill_manual is how we actually split the data
  scale_fill_manual(
    values = c("#FF6A6A", "#87CEFA"),
    name = "Sex",
    breaks = c("Female", "Male"),
    labels = c("Female", "Male")
  )
arch.plot
```

```{r}
# Arch plot comparing vasectomized and non-vasectomized male arching
vas.arch.plot<-ggplot(df.male, aes(treatment, arch, fill = as.factor(contraceptive.status))) +
  geom_split_violin(trim = TRUE) +
  # stat_summary() lets us display a descriptive statistic on a distribution
  stat_summary(fun = mean, fun.min = mean, fun.max = mean,
               geom = "crossbar", 
               width = 0.25,
               position = position_dodge(width = .25),
  ) +
  labs(x = "Treatment", y = "Arch (intervals/trial)") +
  theme_classic() +
  theme(text = element_text(size = 20)) +
  # scale_x_discrete is our different treatments or categories
  scale_x_discrete(labels = c("0" = "Female_Urine", "1" = "Male_Urine", "2" = "Saline_Control")) +
  # scale_fill_manual is how we actually split the data
  scale_fill_manual(
    values = c("#ffffff", "#a6a6a6"),
    name = "Focal Male Status",
    breaks = c("Vasectomized", "None"),
    labels = c("Vasectomized", "None")
  )
vas.arch.plot
```

*Chirp*  
```{r}
chirp.plot<-ggplot(df, aes(treatment, chirp.rounded, fill = as.factor(sex))) +
  geom_split_violin(trim = TRUE) +
  stat_summary(fun = mean, fun.min = mean, fun.max = mean,
               geom = "crossbar", 
               width = 0.25,
               position = position_dodge(width = .25),
  ) +
  labs(x = "Treatment", y = "Chirp (intervals/trial)") +
  theme_classic() +
  theme(text = element_text(size = 20)) +
  scale_x_discrete(labels = c("0" = "Female_Urine", "1" = "Male_Urine", "2" = "Saline_Control")) +
  scale_fill_manual(
    values = c("#FF6A6A", "#87CEFA"),
    name = "Sex",
    breaks = c("Female", "Male"),
    labels = c("Female", "Male")
  )
chirp.plot
```

```{r}
# Chirp plot comparing vasectomized and non-vasectomized male arching
vas.chirp.plot<-ggplot(df.male, aes(treatment, chirp.rounded, fill = as.factor(contraceptive.status))) +
  geom_split_violin(trim = TRUE) +
  # stat_summary() lets us display a descriptive statistic on a distribution
  stat_summary(fun = mean, fun.min = mean, fun.max = mean,
               geom = "crossbar", 
               width = 0.25,
               position = position_dodge(width = .25),
  ) +
  labs(x = "Treatment", y = "Chirp (intervals/trial)") +
  theme_classic() +
  theme(text = element_text(size = 20)) +
  # scale_x_discrete is our different treatments or categories
  scale_x_discrete(labels = c("0" = "Female_Urine", "1" = "Male_Urine", "2" = "Saline_Control")) +
  # scale_fill_manual is how we actually split the data
  scale_fill_manual(
    values = c("#ffffff", "#a6a6a6"),
    name = "Contraceptive Status",
    breaks = c("Vasectomized", "None"),
    labels = c("Vasectomized", "None")
  )
vas.chirp.plot
```

*Urine*  
```{r}
urine.plot<-ggplot(df, aes(treatment, urine.rounded, fill = as.factor(sex))) +
  geom_split_violin(trim = TRUE) +
  stat_summary(fun = mean, fun.min = mean, fun.max = mean,
               geom = "crossbar", 
               width = 0.25,
               position = position_dodge(width = .25),
  ) +
  labs(x = "Treatment", y = "Urine (times/trial)") +
  theme_classic() +
  theme(text = element_text(size = 20)) +
  scale_x_discrete(labels = c("0" = "Female_Urine", "1" = "Male_Urine", "2" = "Saline_Control")) +
  scale_fill_manual(
    values = c("#FF6A6A", "#87CEFA"),
    name = "Sex",
    breaks = c("Female", "Male"),
    labels = c("Female", "Male")
  )
urine.plot
```

```{r}
# Urine plot comparing vasectomized and non-vasectomized male arching
vas.urinate.plot<-ggplot(df.male, aes(treatment, urine.rounded, fill = as.factor(contraceptive.status))) +
  geom_split_violin(trim = TRUE) +
  # stat_summary() lets us display a descriptive statistic on a distribution
  stat_summary(fun = mean, fun.min = mean, fun.max = mean,
               geom = "crossbar", 
               width = 0.25,
               position = position_dodge(width = .25),
  ) +
  labs(x = "Treatment", y = "Urinate (times/trial)") +
  theme_classic() +
  theme(text = element_text(size = 20)) +
  # scale_x_discrete is our different treatments or categories
  scale_x_discrete(labels = c("0" = "Female_Urine", "1" = "Male_Urine", "2" = "Saline_Control")) +
  # scale_fill_manual is how we actually split the data
  scale_fill_manual(
    values = c("#ffffff", "#a6a6a6"),
    name = "Focal Male Status",
    breaks = c("Vasectomized", "None"),
    labels = c("Vasectomized", "None")
  )
vas.urinate.plot
```

*Subcaudal scent mark*  
```{r}
subcsm.plot<-ggplot(df, aes(treatment, subc.sm, fill = as.factor(sex))) +
  geom_split_violin(trim = TRUE) +
  stat_summary(fun = mean, fun.min = mean, fun.max = mean,
               geom = "crossbar", 
               width = 0.25,
               position = position_dodge(width = .25),
  ) +
  labs(x = "Treatment", y = "Subcaudal scent mark (times/trial)") +
  theme_classic() +
  theme(text = element_text(size = 20)) +
  scale_x_discrete(labels = c("0" = "Female_Urine", "1" = "Male_Urine", "2" = "Saline_Control")) +
  scale_fill_manual(
    values = c("#FF6A6A", "#87CEFA"),
    name = "Sex",
    breaks = c("Female", "Male"),
    labels = c("Female", "Male")
  )
subcsm.plot
```

```{r}
# Subc.sm plot comparing vasectomized and non-vasectomized male arching
vas.subcsm.plot<-ggplot(df.male, aes(treatment, subc.sm, fill = as.factor(contraceptive.status))) +
  geom_split_violin(trim = TRUE) +
  # stat_summary() lets us display a descriptive statistic on a distribution
  stat_summary(fun = mean, fun.min = mean, fun.max = mean,
               geom = "crossbar", 
               width = 0.25,
               position = position_dodge(width = .25),
  ) +
  labs(x = "Treatment", y = "Subcaudal scent mark (times/trial)") +
  theme_classic() +
  theme(text = element_text(size = 20)) +
  # scale_x_discrete is our different treatments or categories
  scale_x_discrete(labels = c("0" = "Female_Urine", "1" = "Male_Urine", "2" = "Saline_Control")) +
  # scale_fill_manual is how we actually split the data
  scale_fill_manual(
    values = c("#ffffff", "#a6a6a6"),
    name = "Focal Male Status",
    breaks = c("Vasectomized", "None"),
    labels = c("Vasectomized", "None")
  )
vas.subcsm.plot
```

*Defecate*  
```{r}
defecate.plot<-ggplot(df, aes(treatment, defecate, fill = as.factor(sex))) +
  geom_split_violin(trim = TRUE) +
  stat_summary(fun = mean, fun.min = mean, fun.max = mean,
               geom = "crossbar", 
               width = 0.25,
               position = position_dodge(width = .25),
  ) +
  labs(x = "Treatment", y = "Defecate (times/trial)") +
  theme_classic() +
  theme(text = element_text(size = 20)) +
  scale_x_discrete(labels = c("0" = "Female_Urine", "1" = "Male_Urine", "2" = "Saline_Control")) +
  scale_fill_manual(
    values = c("#FF6A6A", "#87CEFA"),
    name = "Sex",
    breaks = c("Female", "Male"),
    labels = c("Female", "Male")
  )
defecate.plot
```

```{r}
# Defecate plot comparing vasectomized and non-vasectomized male arching
vas.defecate.plot<-ggplot(df.male, aes(treatment, defecate, fill = as.factor(contraceptive.status))) +
  geom_split_violin(trim = TRUE) +
  # stat_summary() lets us display a descriptive statistic on a distribution
  stat_summary(fun = mean, fun.min = mean, fun.max = mean,
               geom = "crossbar", 
               width = 0.25,
               position = position_dodge(width = .25),
  ) +
  labs(x = "Treatment", y = "Defecate (times/trial)") +
  theme_classic() +
  theme(text = element_text(size = 20)) +
  # scale_x_discrete is our different treatments or categories
  scale_x_discrete(labels = c("0" = "Female_Urine", "1" = "Male_Urine", "2" = "Saline_Control")) +
  # scale_fill_manual is how we actually split the data
  scale_fill_manual(
    values = c("#ffffff", "#a6a6a6"),
    name = "Focal Male Status",
    breaks = c("Vasectomized", "None"),
    labels = c("Vasectomized", "None")
  )
vas.defecate.plot
```

