---
title: "aotus.desc.stats"
author: "Isabel Arcusa"
date: "2024-08-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The purpose of this file is to calculate descriptive statistics on response behaviors measured in the Aotus 2021 experiment.  

We will be calculating the median and range for each response behavior: arch, subcaudal scent marking, anogenital sniffing, urine drinking, close proximity, Hinde's index, attempts to mate/mount, and defecating.    

```{r}
# Install dplyr 
library(dplyr)

# Set working directory
setwd("~/Desktop/aotus.fsc.repo")

# Import raw experiment data
raw.df<-read.csv("chemical.mateguarding.data.csv", header = TRUE)

# Eliminate columns that are not needed for these models
# Rename edited data frame as df
df<-raw.df %>% select(-uw.and.agscratch, -prx.23, -anointing, -chuck, -yawn, -lip.smack, 
                      -self.groom, -partner.groom)

# Round chirp and urine columns to the nearest whole number and add as a new column
# We can use mutate() and round() from dplyr() to achieve this
# Add a rounded chirp column
df<-df %>%
  mutate(chirp.rounded = round(df$chirp))

# Add a rounded urine column
df<-df %>%
  mutate(urine.rounded = round(df$urine))

# Some values for prx.01 will have to be rounded
# Add a rounded prx.01 column
df<-df %>%
  mutate(prx.01.rounded = round(df$prx.01))

# Verify that the data frame looks as it should
head(df)
```

Separate sexes.  
```{r}
# Remove rows where sex is "Male"
df.female <- df %>%
  filter(df$sex != "Male")

# Check to make sure the data frame only has female entries
head(df.female)

# Remove rows where sex is "Female"
df.male <- df %>%
  filter(df$sex != "Female")

# Check to make sure the data frame only has male entries
head(df.male)
```

Separate males by whether they are vasectomized or not.  
```{r}
# Remove rows where contraceptive status is "Vasectomized"
df.none.vas <- df.male %>%
  filter(df.male$contraceptive.status != "Vasectomized")

# Check to make sure the data frame only has vasectomized entries
head(df.none.vas)

# Remove rows where contraceptive status is "None"
df.vas <- df.male %>%
  filter(df.male$contraceptive.status != "None")

# Check to make sure the data frame only has none entries
head(df.vas)
```

*Arch*  
```{r}
# Male median
median(df.male$arch)

# Male range
range(df.male$arch)

# Female median
median(df.female$arch)

# Female range
range(df.female$arch)

# Male vas median
median(df.vas$arch)

# Male vas range
range(df.vas$arch)

# Male non.vas median
median(df.none.vas$arch)

# Male non.vas range
range(df.none.vas$arch)
```

*Chirp*  
```{r}
# Male median
median(df.male$chirp.rounded)

# Male range
range(df.male$chirp.rounded)

# Female median
median(df.female$chirp.rounded)

# Female range
range(df.female$chirp.rounded)

# Male vas median
median(df.vas$chirp.rounded)

# Male vas range
range(df.vas$chirp.rounded)

# Male non.vas median
median(df.none.vas$chirp.rounded)

# Male non.vas range
range(df.none.vas$chirp.rounded)
```

*Urine*  
```{r}
# Male median
median(df.male$urine.rounded)

# Male range
range(df.male$urine.rounded)

# Female median
median(df.female$urine.rounded)

# Female range
range(df.female$urine.rounded)

# Male vas median
median(df.vas$urine.rounded)

# Male vas range
range(df.vas$urine.rounded)

# Male non.vas median
median(df.none.vas$urine.rounded)

# Male non.vas range
range(df.none.vas$urine.rounded)
```

```{r}
# Median urinate counts for females
female.median.urine.counts <- df.female %>%
  group_by(treatment) %>%
  summarize(median.urine.count = median(urine.rounded))

# Print the result
print(female.median.urine.counts)

# Range urinate counts for females
female.range.urine.counts <- df.female %>%
  group_by(treatment) %>%
  summarize(range.urine.count = range(urine.rounded))

# Print the result
print(female.range.urine.counts)

# Median urinate counts for males
male.median.urine.counts <- df.male %>%
  group_by(treatment) %>%
  summarize(median.urine.count = median(urine.rounded))

# Print the result
print(male.median.urine.counts)

# Range urinate counts for females
male.range.urine.counts <- df.male %>%
  group_by(treatment) %>%
  summarize(range.urine.count = range(urine.rounded))

# Print the result
print(male.range.urine.counts)
```

*Defecate*  
```{r}
# Male median
median(df.male$defecate)

# Male range
range(df.male$defecate)

# Female median
median(df.female$defecate)

# Female range
range(df.female$defecate)

# Male vas median
median(df.vas$defecate)

# Male vas range
range(df.vas$defecate)

# Male non.vas median
median(df.none.vas$defecate)

# Male non.vas range
range(df.none.vas$defecate)
```

*Subcaudal scent mark*  
```{r}
# Male median
median(df.male$subc.sm)

# Male range
range(df.male$subc.sm)

# Female median
median(df.female$subc.sm)

# Female range
range(df.female$subc.sm)

# Male vas median
median(df.vas$subc.sm)

# Male vas range
range(df.vas$subc.sm)

# Male non.vas median
median(df.none.vas$subc.sm)

# Male non.vas range
range(df.none.vas$subc.sm)
```

```{r}
# Median subc.sm counts for females
female.median.subc.sm.counts <- df.female %>%
  group_by(treatment) %>%
  summarize(median.subc.sm.count = median(subc.sm))

# Print the result
print(female.median.subc.sm.counts)

# Range subc.sm counts for females
female.range.subc.sm.counts <- df.female %>%
  group_by(treatment) %>%
  summarize(range.subc.sm.count = range(subc.sm))

# Print the result
print(female.range.subc.sm.counts)

# Median subc.sm counts for males
male.median.subc.sm.counts <- df.male %>%
  group_by(treatment) %>%
  summarize(median.subc.sm.count = median(subc.sm))

# Print the result
print(male.median.subc.sm.counts)

# Range subc.sm counts for females
male.range.subc.sm.counts <- df.male %>%
  group_by(treatment) %>%
  summarize(range.subc.sm.count = range(subc.sm))

# Print the result
print(male.range.subc.sm.counts)
```

*Hinde's Index*  
```{r}
# Median hindes.index counts for females
female.median.hindes.index.counts <- df.female %>%
  group_by(treatment) %>%
  summarize(median.hindes.index.count = median(hindes.index))

# Print the result
print(female.median.hindes.index.counts)

# Range hindes.index counts for females
female.range.hindes.index.counts <- df.female %>%
  group_by(treatment) %>%
  summarize(range.hindes.index.count = range(hindes.index))

# Print the result
print(female.range.hindes.index.counts)

# Median hindes.index counts for males
male.median.hindes.index.counts <- df.male %>%
  group_by(treatment) %>%
  summarize(median.hindes.index.count = median(hindes.index))

# Print the result
print(male.median.hindes.index.counts)

# Range hindes.index counts for females
male.range.hindes.index.counts <- df.male %>%
  group_by(treatment) %>%
  summarize(range.hindes.index.count = range(hindes.index))

# Print the result
print(male.range.hindes.index.counts)
```

*Close proximity*  
```{r}
# Median prx.01 counts for females
female.median.prx.01.counts <- df.female %>%
  group_by(treatment) %>%
  summarize(median.prx.01.count = median(prx.01))

# Print the result
print(female.median.prx.01.counts)

# Range prx.01 counts for females
female.range.prx.01.counts <- df.female %>%
  group_by(treatment) %>%
  summarize(range.prx.01.count = range(prx.01))

# Print the result
print(female.range.prx.01.counts)

# Median prx.01 counts for males
male.median.prx.01.counts <- df.male %>%
  group_by(treatment) %>%
  summarize(median.prx.01.count = median(prx.01))

# Print the result
print(male.median.prx.01.counts)

# Range prx.01 counts for females
male.range.prx.01.counts <- df.male %>%
  group_by(treatment) %>%
  summarize(range.prx.01.count = range(prx.01))

# Print the result
print(male.range.prx.01.counts)
```
